Indicators for Ecosystem Condition in Norway

### 2023-09-28 alien_species.Rmd

@@

## Introduction
This indicator represent the the ecological on-site effect from  alien species. Whilst the introduction of alien species is best considered a pressure variable, the relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a valid condition indicator.

@@

## About the underlying data

@@

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values

@@

## Analyses

@@

#### Validate

>>>>

Norwegian name: **Fravær av fremmede arter**

@@

## Introduction {#intro-ali}
This indicator represent the the ecological on-site effect from  alien species. Whilst the introduction of alien species is best considered a pressure variable, the relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a valid condition indicator.

@@

## About the underlying data {#about-data-ali}

@@

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-vals-ali}

@@

## Analyses {analyses-ali}

@@

#### Validate {#validate-ali}

### 2023-09-28 functional_plant_indicators_natopen.Rmd

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring programs ANO (Tingstad et al. 2019) and GRUK (Evju et al. 2020) for naturally open ecosystems below tree line (abbreviated as 'nat-open' henceforth). We calculate the functional signature of plant communities in the monitored sites with respect to Grime's CSR values, light, nitrogen, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each nat-open ecosystem type, calculated from 'generalized species lists' developed for ecosystem types in the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the nitrogen signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Framstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

@@

## Analyses
### Data sets

>>>>

## Introduction {#intro-fpin}
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring programs ANO (Tingstad et al. 2019) and GRUK (Evju et al. 2020) for naturally open ecosystems below tree line (abbreviated as 'nat-open' henceforth). We calculate the functional signature of plant communities in the monitored sites with respect to Grime's CSR values, light, nitrogen, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each nat-open ecosystem type, calculated from 'generalized species lists' developed for ecosystem types in the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the nitrogen signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-vals-fpin}
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Framstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

@@

## Analyses {#analyses-fpin}
### Data sets

### 2023-09-28 functional_plant_indicators_wetland.Rmd
Shorten chapter name and list indicator names #77

# Functional plant indicators (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}

@@

@@

@@

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for wetland ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021).

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

@@

## Analyses
### Data sets

@@

### Eksport file (final product)
<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

>>>>

# Functional Plant Indicators - Wetlands {#Functional-plant-indicators-wetland}

@@

Norwegian name: **Planteindikatorer**

@@

Indicators described in this chaper:

- Moisture
- Light
- pH
- Nitrogen

@@

## Introduction {#intro-fpiw}
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for wetland ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021).

@@

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-values-fpiw}
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

@@

## Analyses {#analyses-fpiw}
### Data sets

@@

### Eksport file (final product) {#exp-fpiw}
<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-09-28 gjengroing.Rmd
RenderBook12092023
#75 shorten indicator name

# Vegetation regrowth/encroachment (gjengroing) {#gjengroing}

@@

## Introduction

@@

## About the underlying data

@@

### Representativity in time and space

@@

### Original units

@@

### Temporal coverage

@@

### Additional comments about the dataset

@@

### Reference state

@@

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values

@@

## Uncertainties

@@

## Analyses

### Data sets

@@

#### Regions

@@

### Eksport file (final product)

>>>>

# (PART\*) INDICATORS{-}

# Enchroachment {#gjengroing}
<br />

Norwegian name: **Gjengroing**

_Vegetation regrowth/encroachment_

@@

## Introduction {#intro-enc}

@@

## About the underlying data {#about-data-enc}

@@

### Representativity in time and space {#rep-enc}

@@

### Original units {#units-enc}

@@

### Temporal coverage {#temp-enc}

@@

### Additional comments about the dataset {#comm-enc}

@@

### Reference state {#state-enc}

@@

### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-enc}

@@

## Uncertainties {#err-enc}

@@

## Analyses {#analyses-enc}

### Data sets {#data-enc}

@@

#### Regions {#reg-enc}

@@

### Eksport file (final product) {#exp-enc}

### 2023-09-28 homogeneous_impact_areas.Rmd

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tmap)
library(stars);packageVersion("stars")
library(units)
library(ggpubr)
library(DT)
library(tidyverse)
library(units)
```

## Introduction

This chapter documents the creation of a wall-to-wall map of homogeneous impact areas in Norway.
The map is produced by binning values in the infrastructure index into four discrete categories.
It is likely to be a good predictor for some indicators, such as *slitasje* and the presence of *alien species*.

@@

## Export
I will write this data to the NINA P server.

>>>>

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tmap)
library(stars);packageVersion("stars")
library(units)
library(ggpubr)
library(DT)
#library(tidyverse)
library(ggplot2)
library(dplyr)
library(units)
```

## Introduction {#intro-hia}

This chapter documents the creation of a wall-to-wall map of homogeneous impact areas in Norway.
The map is produced by binning values in the infrastructure index into four discrete categories.
It is likely to be a good predictor for some indicators, such as [*slitasje*](#slitasje) and the presence of [*alien species*](#alien-species).

@@

## Export {#exp-hia}
I will write this data to the NINA P server.

### 2023-09-28 index.Rmd
Fixed libpath issue

Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. Indicators for Ecosystem Condition in Norway. https://ninanor.github.io/ecosystemCondition/index.html.doi [TBA]

## License

>>>>

Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H., Venter, Z., Töpper, J., Speed, J.D.M, Nater, C.R., Evju, M., Åström, J. (2023) Indicators for Ecosystem Condition in Norway. https://doi.org/10.5281/zenodo.7854835

**Contributions**
The list of authors include code developers and reviewers. Not all authors interacted with GitHub through their own user, but for those that did, theiy contributions can be tracked, for example [here](https://github.com/NINAnor/ecosystemCondition/graphs/contributors).

## License {.unnumbered}
